Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate

### 2018-02-26 paper/JAIR/jair.tex

\documentclass{article}
%\usepackage{jair, theapa, rawfonts}

%\usepackage[utf8]{inputenc}
%\usepackage{fontspec}
\usepackage{xeCJK}
%\usepackage{armtex}

\usepackage{multirow}
\usepackage{hyperref}
\usepackage{float}
\usepackage{array}
\usepackage{dirtytalk}
\usepackage{csquotes}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{pifont}

@@

\usepackage[numbers]{natbib}
\bibliographystyle{plainnat}

%\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
%\author{\name Marcelo Prates \email morprates@inf.ufrgs.br \\
%\name Luis Lamb \email lamb@inf.ufrgs.br \\
%\addr Institute of Informatics \\
%Federal University of Rio Grande do Sul \\
%Porto Alegre, 91501-970 Brazil}

\title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}
\author{Marcelo Prates \and Luis Lamb}

@@

@@

\bibliography{main}
\bibliographystyle{named}

>>>>

\documentclass[jair,twoside,11pt,theapa]{article}
\usepackage[bookmarks=true]{hyperref}
\hypersetup{%
    bookmarks=false,                            % show bookmarks bar?
    pdftitle={TeX and friends hackers manual},  % title
    pdfauthor={Yiannis Lazarides},              % author
    pdfsubject={TeX and LaTeX},                 % subject of the document
    pdfkeywords={TeX, LaTeX, graphics, images}, % list of keywords
    %colorlinks=true,                            % false: boxed links; true: colored links
    %linkcolor=blue,                             % color of internal links
    %citecolor=black,                            % color of links to bibliography
    %filecolor=black,                            % color of file links
    %urlcolor=purple,                            % color of external links
    linktoc=page                                % only page is linked
}
\usepackage{jair, theapa, rawfonts}

\usepackage[utf8]{inputenc}
\usepackage{xeCJK}
\usepackage{float}
\usepackage{graphicx}
\usepackage{dirtytalk}
\usepackage{amsmath}
\usepackage{array}
\usepackage{csquotes}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{multirow}

@@

\jairheading{1}{2018}{1-15}{?/2017}{?/2018}
\ShortHeadings{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
{Prates \& Lamb}
\firstpageno{25}

@@

\title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}

\author{\name Marcelo Prates \email morprates@inf.ufrgs.br \\
       \name Luis Lamb \email lamb@inf.ufrgs.br \\
       \addr Institute of Informatics \\
Federal University of Rio Grande do Sul \\
Porto Alegre, 91501-970 Brazil}

% For research notes, remove the comment character in the line below.
% \researchnote

@@

\vskip 0.2in

\bibliographystyle{theapa}
\bibliography{jair}

### 2018-02-26 paper/Palgrave/main.tex

\documentclass[fleqn,10pt]{wlscirep}

\usepackage{multirow}
\usepackage{hyperref}
\usepackage{float}
\usepackage{array}
\usepackage{dirtytalk}
\usepackage{csquotes}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{pifont}

@@

%\usepackage[numbers]{natbib}
%\bibliographystyle{plainnat}

@@

\author[1,*]{Marcelo Prates}
\author[1,+]{Luis Lamb}
\affil[1]{Federal University of Rio Grande do Sul, Institute of Informatics, Porto Alegre, 91501-970, Brazil}

\affil[*]{morprates@inf.ufrgs.br}
\affil[+]{lamb@inf.ufrgs.br}

%\keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
Recently, there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of bias and prejudices. A large number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI. 
In this paper, we construct sentences with gender neutral pronouns in different languages 
and translate these sentences into English using the Google Translate API. 
The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate 
API computes for each sentence. We show that not only there is a tendency towards translating pronouns to
masculine, but also that this tendency is exaggerated for sentences about male dominated fields such as Computer Science. On the other hand, sentences about artistic fields are less likely to yield male defaults. This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of the people behind their training datasets -- ourselves.
\end{abstract}
\begin{document}

\flushbottom
\maketitle
% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
%
%  Click the title above to edit the author information and abstract
%
\thispagestyle{empty}

@@

\begin{thebibliography}{10}
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{\texttt{#1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
\expandafter\ifx\csname doiprefix\endcsname\relax\def\doiprefix{DOI }\fi
\providecommand{\bibinfo}[2]{#2}
\providecommand{\eprint}[2][]{\url{#2}}

\bibitem{dascal1982universal}
\bibinfo{author}{Dascal, M.}
\newblock \bibinfo{journal}{\bibinfo{title}{Universal language schemes in
  {E}ngland and {F}rance, 1600-1800 comments on {J}ames {K}nowlson}}.
\newblock {\emph{\JournalTitle{Studia leibnitiana}}}
  \textbf{\bibinfo{volume}{14}}, \bibinfo{pages}{98--109}
  (\bibinfo{year}{1982}).

\bibitem{locke1955machine}
\bibinfo{author}{Locke, W.~N.} \& \bibinfo{author}{Booth, A.~D.}
\newblock \emph{\bibinfo{title}{Machine translation of languages: fourteen
  essays}} (\bibinfo{publisher}{Published jointly by Technology Press of the
  Massachusetts Institute of Technology and Wiley, New York},
  \bibinfo{year}{1955}).

\bibitem{weaver1955translation}
\bibinfo{author}{Weaver, W.}
\newblock \bibinfo{title}{Translation}.
\newblock In \bibinfo{editor}{Locke, W.~N.} \& \bibinfo{editor}{Booth, A.~D.}
  (eds.) \emph{\bibinfo{booktitle}{Machine translation of languages}},
  vol.~\bibinfo{volume}{14}, \bibinfo{pages}{15--23}
  (\bibinfo{publisher}{Cambridge: Technology Press, MIT},
  \bibinfo{year}{1955}).
\newblock \urlprefix\url{http://www.mt-archive.info/Weaver-1949.pdf}.
\newblock \bibinfo{note}{Last visited 2017-12-17}.

\bibitem{gordin2015scientific}
\bibinfo{author}{Gordin, M.~D.}
\newblock \emph{\bibinfo{title}{Scientific Babel: How science was done before
  and after global English}} (\bibinfo{publisher}{University of Chicago Press},
  \bibinfo{year}{2015}).

\bibitem{hutchins1986machine}
\bibinfo{author}{Hutchins, W.~J.}
\newblock \emph{\bibinfo{title}{Machine translation: past, present, future}}
  (\bibinfo{publisher}{Ellis Horwood Chichester}, \bibinfo{year}{1986}).

\bibitem{koehn2009statistical}
\bibinfo{author}{Koehn, P.}
\newblock \emph{\bibinfo{title}{Statistical machine translation}}
  (\bibinfo{publisher}{Cambridge University Press}, \bibinfo{year}{2009}).

\bibitem{Moses2007}
\bibinfo{author}{Koehn, P.} \emph{et~al.}
\newblock \bibinfo{title}{Moses: Open source toolkit for statistical machine
  translation}.
\newblock In \emph{\bibinfo{booktitle}{{ACL} 2007, Proceedings of the 45th
  Annual Meeting of the Association for Computational Linguistics, June 23-30,
  2007, Prague, Czech Republic}} (\bibinfo{year}{2007}).
\newblock \urlprefix\url{http://aclweb.org/anthology/P07-2045}.

\bibitem{Bahdanau2014}
\bibinfo{author}{Bahdanau, D.}, \bibinfo{author}{Cho, K.} \&
  \bibinfo{author}{Bengio, Y.}
\newblock \bibinfo{journal}{\bibinfo{title}{Neural machine translation by
  jointly learning to align and translate}}.
\newblock {\emph{\JournalTitle{CoRR}}} \textbf{\bibinfo{volume}{abs/1409.0473}}
  (\bibinfo{year}{2014}).
\newblock \urlprefix\url{http://arxiv.org/abs/1409.0473}.
\newblock \eprint{1409.0473}.

\bibitem{carl2003recent}
\bibinfo{author}{Carl, M.} \& \bibinfo{author}{Way, A.}
\newblock \emph{\bibinfo{title}{Recent advances in example-based machine
  translation}}, vol.~\bibinfo{volume}{21} (\bibinfo{publisher}{Springer
  Science \& Business Media}, \bibinfo{year}{2003}).

\bibitem{Firat2017}
\bibinfo{author}{Firat, O.}, \bibinfo{author}{Cho, K.},
  \bibinfo{author}{Sankaran, B.}, \bibinfo{author}{Yarman{-}Vural, F.~T.} \&
  \bibinfo{author}{Bengio, Y.}
\newblock \bibinfo{journal}{\bibinfo{title}{Multi-way, multilingual neural
  machine translation}}.
\newblock {\emph{\JournalTitle{Computer Speech {\&} Language}}}
  \textbf{\bibinfo{volume}{45}}, \bibinfo{pages}{236--252}
  (\bibinfo{year}{2017}).
\newblock \urlprefix\url{https://doi.org/10.1016/j.csl.2016.10.006}.
\newblock \doiprefix 10.1016/j.csl.2016.10.006.

\bibitem{wu2016google}
\bibinfo{author}{Johnson, M.} \emph{et~al.}
\newblock \bibinfo{journal}{\bibinfo{title}{Google's multilingual neural
  machine translation system: Enabling zero-shot translation}}.
\newblock {\emph{\JournalTitle{{TACL}}}} \textbf{\bibinfo{volume}{5}},
  \bibinfo{pages}{339--351} (\bibinfo{year}{2017}).
\newblock
  \urlprefix\url{https://transacl.org/ojs/index.php/tacl/article/view/1081}.

\bibitem{Chomsky2011}
\bibinfo{author}{Chomsky, N.}
\newblock \bibinfo{title}{The golden age: A look at the original roots of
  artificial intelligence, cognitive science, and neuroscience (partial
  transcript of an interview with {N}. {C}homsky at {MIT150} {S}ymposia:
  Brains, minds and machines symposium} (\bibinfo{year}{2011}).
\newblock \urlprefix\url{https://chomsky.info/20110616/}.
\newblock \bibinfo{note}{Last visited 2017-12-26}.

\bibitem{norvig2017chomsky}
\bibinfo{author}{Norvig, P.}
\newblock \bibinfo{title}{On {C}homsky and the two cultures of statistical
  learning} (\bibinfo{year}{2017}).
\newblock \urlprefix\url{http://norvig.com/chomsky.html}.
\newblock \bibinfo{note}{Last visited 2017-12-17}.

\bibitem{schiebinger2014scientific}
\bibinfo{author}{Schiebinger, L.}
\newblock \bibinfo{journal}{\bibinfo{title}{Scientific research must take
  gender into account}}.
\newblock {\emph{\JournalTitle{Nature}}} \textbf{\bibinfo{volume}{507}},
  \bibinfo{pages}{9} (\bibinfo{year}{2014}).

\bibitem{garcia2016racist}
\bibinfo{author}{Garcia, M.}
\newblock \bibinfo{journal}{\bibinfo{title}{Racist in the machine: The
  disturbing implications of algorithmic bias}}.
\newblock {\emph{\JournalTitle{World Policy Journal}}}
  \textbf{\bibinfo{volume}{33}}, \bibinfo{pages}{111--117}
  (\bibinfo{year}{2016}).

\bibitem{angwin2016machine}
\bibinfo{author}{Angwin, J.}, \bibinfo{author}{Larson, J.},
  \bibinfo{author}{Mattu, S.} \& \bibinfo{author}{Kirchner, L.}
\newblock \bibinfo{title}{Machine bias: {T}here's software used across the
  country to predict future criminals and it's biased against blacks}
  (\bibinfo{year}{2016}).
\newblock \urlprefix\url{https://www.
  propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}.
\newblock \bibinfo{note}{Last visited 2017-12-17}.

\bibitem{wang2017deep}
\bibinfo{author}{Wang, Y.} \& \bibinfo{author}{Kosinski, M.}
\newblock \bibinfo{title}{Deep neural networks are more accurate than humans at
  detecting sexual orientation from facial images} (\bibinfo{year}{2017}).

\bibitem{womanunlockphone2017}
\bibinfo{author}{Papenfuss, M.}
\newblock \bibinfo{title}{Woman in {C}hina says colleague's face was able to
  unlock her i{P}hone {X}} (\bibinfo{year}{2017}).
\newblock
  \urlprefix\url{http://www.huffpostbrasil.com/entry/iphone-face-recognition-double_us_5a332cbce4b0ff955ad17d50}.
\newblock \bibinfo{note}{Last visited 2017-12-17}.

\bibitem{racistsoapdispenser2017}
\bibinfo{author}{Mills, K.-A.}
\newblock \bibinfo{title}{'{R}acist' soap dispenser refuses to help
  dark-skinned man wash his hands - but {T}witter blames 'technology'}
  (\bibinfo{year}{2017}).
\newblock
  \urlprefix\url{http://www.mirror.co.uk/news/world-news/racist-soap-dispenser-refuses-help-11004385}.
\newblock \bibinfo{note}{Last visited 2017-12-17}.

\end{thebibliography}

%\bibliography{main}
%\bibliographystyle{named}

\end{document}

>>>>

\documentclass[fleqn,10pt]{article}

\usepackage{multirow}
\usepackage{float}
\usepackage{array}
\usepackage{dirtytalk}
\usepackage{csquotes}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{pifont}

\usepackage{natbib}
\bibliographystyle{agsm}
\let\hardvardurl\url
\usepackage{url}
\usepackage[breaklinks]{hyperref}

@@

@@

\begin{document}

\maketitle

\begin{abstract}
Recently, there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of bias and prejudices. A large number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI. 
In this paper, we construct sentences with gender neutral pronouns in different languages 
and translate these sentences into English using the Google Translate API. 
The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate 
API computes for each sentence. We show that not only there is a tendency towards translating pronouns to
masculine, but also that this tendency is exaggerated for sentences about male dominated fields such as Computer Science. On the other hand, sentences about artistic fields are less likely to yield male defaults. This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of the people behind their training datasets -- ourselves.
\end{abstract}

@@

\begin{thebibliography}{xx}

\harvarditem[Angwin et~al.]{Angwin, Larson, Mattu \harvardand\
  Kirchner}{2016}{angwin2016machine}
Angwin, J., Larson, J., Mattu, S. \harvardand\ Kirchner, L.  \harvardyearleft
  2016\harvardyearright , `Machine bias: {T}here's software used across the
  country to predict future criminals and it's biased against blacks'.
\newblock Last visited 2017-12-17.
\newline\harvardurl{https://www.
  propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}

\harvarditem[Bahdanau et~al.]{Bahdanau, Cho \harvardand\
  Bengio}{2014}{Bahdanau2014}
Bahdanau, D., Cho, K. \harvardand\ Bengio, Y.  \harvardyearleft
  2014\harvardyearright , `Neural machine translation by jointly learning to
  align and translate', {\em CoRR} {\bf abs/1409.0473}.
\newline\harvardurl{http://arxiv.org/abs/1409.0473}

\harvarditem{Carl \harvardand\ Way}{2003}{carl2003recent}
Carl, M. \harvardand\ Way, A.  \harvardyearleft 2003\harvardyearright , {\em
  Recent advances in example-based machine translation}, Vol.~21, Springer
  Science \& Business Media.

\harvarditem{Chomsky}{2011}{Chomsky2011}
Chomsky, N.  \harvardyearleft 2011\harvardyearright , `The golden age: A look
  at the original roots of artificial intelligence, cognitive science, and
  neuroscience (partial transcript of an interview with {N}. {C}homsky at
  {MIT150} {S}ymposia: Brains, minds and machines symposium'.
\newblock Last visited 2017-12-26.
\newline\harvardurl{https://chomsky.info/20110616/}

\harvarditem{Dascal}{1982}{dascal1982universal}
Dascal, M.  \harvardyearleft 1982\harvardyearright , `Universal language
  schemes in {E}ngland and {F}rance, 1600-1800 comments on {J}ames {K}nowlson',
  {\em Studia leibnitiana} {\bf 14}(1),~98--109.

\harvarditem[Firat et~al.]{Firat, Cho, Sankaran, Yarman{-}Vural \harvardand\
  Bengio}{2017}{Firat2017}
Firat, O., Cho, K., Sankaran, B., Yarman{-}Vural, F.~T. \harvardand\ Bengio, Y.
   \harvardyearleft 2017\harvardyearright , `Multi-way, multilingual neural
  machine translation', {\em Computer Speech {\&} Language} {\bf 45},~236--252.
\newline\harvardurl{https://doi.org/10.1016/j.csl.2016.10.006}

\harvarditem{Garcia}{2016}{garcia2016racist}
Garcia, M.  \harvardyearleft 2016\harvardyearright , `Racist in the machine:
  The disturbing implications of algorithmic bias', {\em World Policy Journal}
  {\bf 33}(4),~111--117.

\harvarditem{Gordin}{2015}{gordin2015scientific}
Gordin, M.~D.  \harvardyearleft 2015\harvardyearright , {\em Scientific Babel:
  How science was done before and after global English}, University of Chicago
  Press.

\harvarditem{Hutchins}{1986}{hutchins1986machine}
Hutchins, W.~J.  \harvardyearleft 1986\harvardyearright , {\em Machine
  translation: past, present, future}, Ellis Horwood Chichester.

\harvarditem[Johnson et~al.]{Johnson, Schuster, Le, Krikun, Wu, Chen, Thorat,
  Vi{\'{e}}gas, Wattenberg, Corrado, Hughes \harvardand\
  Dean}{2017}{wu2016google}
Johnson, M., Schuster, M., Le, Q.~V., Krikun, M., Wu, Y., Chen, Z., Thorat, N.,
  Vi{\'{e}}gas, F.~B., Wattenberg, M., Corrado, G., Hughes, M. \harvardand\
  Dean, J.  \harvardyearleft 2017\harvardyearright , `Google's multilingual
  neural machine translation system: Enabling zero-shot translation', {\em
  {TACL}} {\bf 5},~339--351.
\newline\harvardurl{https://transacl.org/ojs/index.php/tacl/article/view/1081}

\harvarditem{Koehn}{2009}{koehn2009statistical}
Koehn, P.  \harvardyearleft 2009\harvardyearright , {\em Statistical machine
  translation}, Cambridge University Press.

\harvarditem[Koehn et~al.]{Koehn, Hoang, Birch, Callison{-}Burch, Federico,
  Bertoldi, Cowan, Shen, Moran, Zens, Dyer, Bojar, Constantin \harvardand\
  Herbst}{2007}{Moses2007}
Koehn, P., Hoang, H., Birch, A., Callison{-}Burch, C., Federico, M., Bertoldi,
  N., Cowan, B., Shen, W., Moran, C., Zens, R., Dyer, C., Bojar, O.,
  Constantin, A. \harvardand\ Herbst, E.  \harvardyearleft
  2007\harvardyearright , Moses: Open source toolkit for statistical machine
  translation, {\em in} `{ACL} 2007, Proceedings of the 45th Annual Meeting of
  the Association for Computational Linguistics, June 23-30, 2007, Prague,
  Czech Republic'.
\newline\harvardurl{http://aclweb.org/anthology/P07-2045}

\harvarditem{Locke \harvardand\ Booth}{1955}{locke1955machine}
Locke, W.~N. \harvardand\ Booth, A.~D.  \harvardyearleft 1955\harvardyearright
  , {\em Machine translation of languages: fourteen essays}, Published jointly
  by Technology Press of the Massachusetts Institute of Technology and Wiley,
  New York.

\harvarditem{Mills}{2017}{racistsoapdispenser2017}
Mills, K.-A.  \harvardyearleft 2017\harvardyearright , `'{R}acist' soap
  dispenser refuses to help dark-skinned man wash his hands - but {T}witter
  blames 'technology''.
\newblock Last visited 2017-12-17.
\newline\harvardurl{http://www.mirror.co.uk/news/world-news/racist-soap-dispenser-refuses-help-11004385}

\harvarditem{Norvig}{2017}{norvig2017chomsky}
Norvig, P.  \harvardyearleft 2017\harvardyearright , `On {C}homsky and the two
  cultures of statistical learning'.
\newblock Last visited 2017-12-17.
\newline\harvardurl{http://norvig.com/chomsky.html}

\harvarditem{Papenfuss}{2017}{womanunlockphone2017}
Papenfuss, M.  \harvardyearleft 2017\harvardyearright , `Woman in {C}hina says
  colleague's face was able to unlock her i{P}hone {X}'.
\newblock Last visited 2017-12-17.
\newline\url{http://www.huffpostbrasil.com/entry/iphone-face-recognition-double_us_5a332cbce4b0ff955ad17d50}

\harvarditem{Schiebinger}{2014}{schiebinger2014scientific}
Schiebinger, L.  \harvardyearleft 2014\harvardyearright , `Scientific research
  must take gender into account', {\em Nature} {\bf 507}(7490),~9.

\harvarditem{Wang \harvardand\ Kosinski}{2017}{wang2017deep}
Wang, Y. \harvardand\ Kosinski, M.  \harvardyearleft 2017\harvardyearright ,
  `Deep neural networks are more accurate than humans at detecting sexual
  orientation from facial images'.

\harvarditem{Weaver}{1955}{weaver1955translation}
Weaver, W.  \harvardyearleft 1955\harvardyearright , Translation, {\em in}
  W.~N. Locke \harvardand\ A.~D. Booth, eds, `Machine translation of
  languages', Vol.~14, Cambridge: Technology Press, MIT, pp.~15--23.
\newblock Last visited 2017-12-17.
\newline\harvardurl{http://www.mt-archive.info/Weaver-1949.pdf}

\end{thebibliography}
\end{document}
